### 2015-07-23 RITA2015/OldVersion/intro.tex
New folder for RITA submission.

\documentclass[12pt,twoside]{article}
\usepackage{Jcst}
%Equations,theorems,lemmas,etc. are created using the traditional enviroment
%

\usepackage{xspace}

\usepackage{amsmath}
\usepackage[thmmarks,amsmath]{ntheorem}

\newcommand{\openbox}{\leavevmode
  \hbox to.77778em{%
  \hfil\vrule
  \vbox to.675em{\hrule width.6em\vfil\hrule}%
  \vrule\hfil}}

\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{}
\theoremindent0cm
\theoremnumbering{arabic}
\newtheorem{algo}{Algorithm}

\theoremstyle{plain}
%\theoremheaderfont{\normalfont\itshape}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{}
\theoremindent0cm
\theoremnumbering{arabic}
\theoremsymbol{\ensuremath{\openbox}} 
\newtheorem{example}{Example}

\newcommand{\pisodm}[0]{$\pi$SOD-M\xspace}
\def\FlyingPig{\textsl{FlyingPig}\xspace}

\newcounter{numberInTrivlist}

\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})}
                                         {\usecounter{numberInTrivlist}
                                          \setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ }
                                         {\setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\begin{document}

\setcounter{page}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%Set Page Head%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setpageinformation
%{Head of the first page} {Running head of odd pages}
{}
{ }{ }{}{Apr.}{2015}
{\pisodm: Building Service-Oriented Applications}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%===========================================================

%\begin{CJK}{GBK}{song}

\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}

%\author{Khalid Belhajjame$^1$, Valeria~de~Castro$^2$, Umberto~Souza~da~Costa$^3$, Javier~A.~Espinosa-Oviedo$^4$, Martin~A.~Musicante$^{3,*}$, Pl\'acido~A.~Souza~Neto$^5$, Genoveva~Vargas-Solar$^6$, \\ Jos\'e-Luis~Zechinelli-Martini$^4$}

%\footnotetext{$^*$.Corresponding author.}
%\footnotetext{$^1$.Universit\'e de Paris - Dauphine -- Paris, France.
%\textsl{kbelhajj@googlemail.com}.
%}
%\footnotetext{$^2$.Universidad Rey Juan Carlos -- M\'{o}stoles, Spain.
%\textsl{valeria.decastro@urjc.es}.
%}
%\footnotetext{$^3$.Universidade Federal do Rio Grande do Norte -- Natal, Brazil.
%\textsl{\{umberto,mam\}@dimap.ufrn.br}.
%}
%\footnotetext{$^4$.Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico.
%\textsl{javiera.espinosa@gmail.com}, \textsl{joseluis.zechinelli@udlap.mx}.
%}
%\footnotetext{$^5$.Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil.
%\textsl{placido.neto@ifrn.edu.br}.
%}
%\footnotetext{$^6$.CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France.
%\textsl{genoveva.vargas@imag.fr}.
%}

%\maketitle

\begin{abstract}
Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
The late inclusion of NFRs in the development may compromise the quality of the deployed application.
This paper presents \pisodm, a method and associated tools that
\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
\end{abstract}

\keywords{MDA , Non-Functional Requirements, Service-based software process}

\begin{multicols}{2}
\normalsize

%% main text
%*********************************************************************************************************
\section{Introduction}
\label{sec:intro}

@@

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional}$^7$\footnote{$^7$.Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

@@

The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}$^8$\footnote{$^8$.The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.

@@

The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System$^9$\footnote{$^9$.The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}. In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.

@@

\section{Related Work}
\label{sec:relworks}
While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
NFRs include aspects such as performance, authentication and quality constraints.
These requirements are usually specified by conditions, called non-functional properties.
Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.

Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.

In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
\textit{service level}, \textit{system level} and \textit{business level}.
%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}. The work also defines a \textit{web service quality model}, which considers non-functional properties.
In~\cite{XiaoCZBOLH08} the authors use the terms
\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
The notion of non-functional attribute is used to describe NFRs of the abstract process model.
In the lower level, the composition is annotated with non-functional attributes.

D'Ambrogio~\cite{DAmbrogio06} uses the term \textit{quality category} to group similar \textit{quality characteristics}.
\textit{Quality dimensions} are used to quantify an individual characteristic.
For instance, the quality category \textit{performance} groups characteristics such as
\textit{latency} and \textit{throughput}.
The development process is based on MDA and the authors also present a WSDL extension for describing the QoS of web services. A catalog of \textit{QoS characteristics} is provided for the web service domain, including properties such as \textit{availability}, \textit{reliability} and \textit{access control}.

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

Pastrana et al.~\cite{PastranaPK11} use a traditional design-by-contract approach~\cite{Meyer97}.
The authors use the term \textit{contract} to describe non-functional re\-quire\-ments.
Contracts may present pre-con\-di\-tions, post-conditions and invariants.
Also, a con\-tract may define \textit{assertions} associated with \textit{qual\-i\-ty properties}.
Each service may have as many associated \textit{contracts} as needed.

Chollet et al.~\cite{CholletL09} associate (non-func\-tion\-al) \textit{quality properties} to
(functional) activities.
They present a security meta-model that takes into account web service compositions.
In this work the non-functional requirements considered are  \textit{authentication}, \textit{integrity} and \textit{confidentiality}.
Each NFR is associated with a service activity.

Ceri et al.\cite{CeriDMF07} use the notions of \textit{policy}, \textit{rule}, \textit{condition} and \textit{action model} to specify NFRs.
Agarwal et al.~\cite{AgarwalLS09} associate \textit{service policies} to services.
Each service may also have \textit{properties}, such as \textit{security} and \textit{reliability}.
Ovaska et al.~\cite{OvaskaEHPA10} use the terms \textit{quality attribute}, \textit{category}, \textit{conceptual layer} and \textit{importance} to organize and classify NFRs.
Other authors do not define specific terms to refer to NFRs; they use terms such as \textit{attribute}~\cite{ZhangPSP05,BasinDL06,JeongCL09},
\textit{property}~\cite{Fabra2011},
\textit{factor}~\cite{MohantyRP10,GutierrezRF10},
\textit{characteristic}~\cite{DiamadopoulouMPS08},
\textit{qual\-i\-ty level} \cite{ModicaTV09}, and
\textit{value}~\cite{ThissenW06,BasinDL06}.

Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.

Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}. As Ameller et al. realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}. Those authors also remark the lack of industrial use cases~\cite{Ameller201542}. 
Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.

%Fabra \textit{et al.}~\cite{Fabra2011} also describes the importance of  MDD for service-oriented applications.
%The work  presents a development methodology, although this methodology is not centered on NFRs.
The authors in~\cite{ThissenW06,ZhangPSP05} use formal methods to define a service-based development process that takes NFRs into account.
In~\cite{AgarwalLS09,PastranaPK11} ontologies are used to define and model NFRs,
as  in~\cite{XiaoCZBOLH08,GutierrezRF10} Business Process Modeling is used for
system specification, including NFRs.
In the method defined in~\cite{XiaoCZBOLH08}, each task and data item of the application can be
annotated with functional as well as non-functional attributes (NFAs).
Functional and non-functional at\-tri\-butes are independently defined.
They are attached to specific tasks later in the development of the application.
NFAs for data considers \textit{value} and \textit{range}, whereas NFAs for tasks include \textit{cost}, \textit{time}, \textit{resources} and \textit{expressions}.

The proposal in~\cite{ThissenW06} presents steps for  selecting services
by taking QoS information into account. The proposed steps are:
\textit{(i)} identification of relevant QoS information;
\textit{(ii)} identification of basic composition patterns and
QoS aggregation rules for these patterns; and
\textit{(iii)} definition of a selection mechanism of services.
The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
\textit{availability}.

Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.

Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.

Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.

%\bigskip
%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
%It may use web services provided by third party sources.
%The System level deals with the coordination of services, to implement the business logic.
%Our proposal, shown in the next section, follows these ideas.

\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}

In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.

In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
The presentation is exemplified by a running example.
The transformations between models are presented in section~\ref{sec:mmrules}.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{piSOD-M_process.pdf}
\caption{\pisodm Overview.}
\label{fig:piSOD-M}
\end{figure*}

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Computation Independent Models}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

This level focuses on the highest-level view of the system, including its business and requirement specifications.
At this stage of the development, the structure and system processing details are still unknown or undetermined.
\pisodm uses the \textit{e$^3$value}~\cite{Gordijn02valuebased} and \textit{BPMN}~\cite{BPMN} meta-models for this purpose.

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{E$^3$value meta-model}
% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The e$^3$value model of an application identifies the value/information exchange between components of the system.
It is a business model that represents a business case graphically as a set of value exchanges$^{10}$\footnote{$^{10}$.See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
The model is well suited to enhance the understanding of the environment in which an application is being developed.
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

\begin{figure*}
\center
\includegraphics[width=0.75\textwidth]{e3value.pdf}
\hspace*{5cm}\includegraphics[width=0.4\textwidth]{3ValueKey.pdf}
\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
\end{figure*}

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
Let us consider the scenario ``To Publish Music'', used as a running example in this section:
An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
The following (external) services will be used by the application:
\begin{itemize}
\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
\begin{itemize}
%\item{\sf\small get-Status ( usedid ): String ;
\item {\sf\small update-Status ( userid, new-status ): String};
\end{itemize}
\end{itemize}

The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
The e$^3$value value model shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status.
We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
\end{example}

The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
In order to face this shortcoming, our method proposes the BPMN meta-model as a tool for modelling this aspect of the application.

\subsubsection{BPMN meta-model}

BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow.
The next example illustrates the use of this meta-model.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model$^{11}$\footnote{$^{11}$.Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\end{example}
%
\begin{figure}
\center
\includegraphics[width=0.45\textwidth]{SC.pdf}
\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
\end{figure}

The CIM level models are the basis for the development of the application.
The information presented at this level is (manually) refined into PIM-level models of the method \pisodm described next.
Notice that, at this level, the models are (informally) used to describe both functional and non-functional requirements.

%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -
\subsection{Platform Independent Models}
%. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -. . - -

This level focuses on the system functionality, hiding the details of any particular platform.
The specification defines those parts of the system that do not change from one platform to another.
Our method defines three PIM-level meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-Ser\-vice\-Pro\-cess} and \textit{$\pi$-ServiceComposition}.

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-UseCase} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

Figure~\ref{fig:CIM:usecasemetamodel} presents the \textit{$\pi$-UseCase} meta-model.
The goal of \textit{$\pi$-UseCase}  is to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
The notion of \textit{policy} is used to describe NFRs.
(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe non-functional requirements with  the following  concepts:  {\sc Business service}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
An {\sc End Consumer} is represented by an {\sc Actor}.
An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), while a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
 The {\sc Business Collaborator} concept is represented  through {\sc Packages}.
A {\sc Business Collaborator} represents an external service or a system that interacts with the application that are being modeled.
Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
%Thus, the services functions can be specified as being grouped into packages.

 \begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{UseCaseMetaModel.pdf}
\caption{\label{fig:CIM:usecasemetamodel} $\pi$-UseCase Meta-Model.}
\end{figure*}

 The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint}) use case.
There are three kinds of constraint:
\textit{(i)} A restriction may be associated to data ({\sc Value Constraint}), represented as the stereotype ``value'';
\textit{(ii)} A restriction may correspond to business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
\textit{(iii)} A restriction may describe {\sc Exceptional Behavior} constraints, represented as the stereotype {\sf ``exceptional\_behavior''}.
The use of these constraints is shown in the following example.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example application.
We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the (i) conditions imposed by services usage --for example, the fact that both Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).

The ``To Publish Music'' Business Service expects  the Facebook or Twitter user status to be changed every time a user starts listening a new song.
Therefore, it is necessary to perform a social network authentication with the users data. Each social network uses different services and different forms of authentication. The authentication constraint is required to update a music status. The restriction is stereotyped as a value constraint, because the users id and password are verified.  Figure \ref{fig:CIM:piusecasetpm} shows the buy music, download music, listen music and pay use cases. The process of buying a song requires the user private data for a Spotify account, and also a secure connection, represented as value and business constraint, respectively. For payment, the user must provide the data of payment card or PayPal account login and password, represented as value stereotype.
Another restriction is that the minimum payment value is 2 euros.
%{\color{red} Talk about this model for the example.}
\end{example}

\begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{UseCase.pdf}
\caption{\label{fig:CIM:piusecasetpm} $\pi$-UseCase model for ``To Publish Music''.}
\end{figure*}

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-ServiceProcess} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The \textit{$\pi$-ServiceProcess} meta-model extends the UML activity diagram with the concept of \textit{contract} to represent constraints over data and actions. This concept is used to model
groups  of  constraints  in the \textit{$\pi$-UseCase}.
In Figure~\ref{fig:CIM:serviceprocessmetamodel}, the concepts of the \textit{$\pi$-Ser\-vice\-Process} meta-model are: {\sc Contract}, {\sc Assertion}, {\sc Exceptional behavior}, {\sc Activity}, {\sc Ser\-vice Acti\-vi\-ty}, {\sc Action} and {\sc Constraint}.
The dotted part of the figure defines those concepts related to the representation of NFRs.

A specific \textit{$\pi$-ServiceProcess} model shows the set of logically related activities to be performed in a service-based application.
So, the activities of this model represent a behavior that is part of the business logic of the application.
A \textit{$\pi$-ServiceProcess}  model contains three main elements: (i) service process, (ii) service activity and (iii) activity contract. A service activity represents an operation that is part of the execution flow, and it is modeled as an {\sc Action}.
An activity contract represents the {\sc Non-functional Requirement}  that is also part of the execution flow of a service, identified  as a stereotyped activity ({\sf assertion}).
The {\sc Assertion}s associated to an {\sc Action} compose a {\sc Contract}. 
Using the concepts {\sc Contract} and {\sc Assertion} it is possible to specify each activity service, by defining its pre-conditions and post-conditions.

\begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{ServiceProcessMetaModel.pdf}
\caption{\label{fig:CIM:serviceprocessmetamodel} $\pi$-Service Process Meta-Model.}
\end{figure*}

%According to the meta-model, a \textit{$\pi$-ServiceProcess} model represents the service processes and its contracts. It models the set of system activities and describe contracts for each activity or the process as a whole. Each service identified in the previous model (\textit{$\pi$-UseCase}) is mapped to an action. This model defines a workflow for  modeling business logic and the behaviour of the application being developed. The behaviour is represented by each action and the restrictions by the assertion description. All assertions are described by stereotyped actions.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic3}
Considering the example scenario, the contract based process of activities  is shown in figure \ref{fig:CIM:serviceprocess}. The buy music and publish music services (update Twitter and Facebook) have pre- and post-conditions assertions that are composed into a contract for each service. The buy music pre-conditions consist in verifying:
(i) if the User data are correct;
(ii) if the User is already logged in Spotify;
(iii) if bank account information are correct and;
(iv) if there are enough funds in the bank account to cover the payment.
A post-condition ensures the complete transaction and verifies if a notification was sent to the user and Spotify, about the payment authorization. There are four assertions for the buy music action, and each assertion has been detailed with the assertion property and predicate that must be verified. To update services, depending of each service, there may be different restrictions. As an example, a new verification of user data and message format is appropriate (maximum 140 characters), in the case of Twitter. In the case of Facebook, it is required that the user is already logged in Spotify and these data are the same as Facebook.
As post-condition, the application ensures that the Facebook service sends a notification of success.
To update Twitter a pre-condition is required, while to update Facebook it is necessary to check a pre-condition and a confirmation notice (modeled as post-condition).
As a pre-condition for ``twitter update'' it is necessary that (i) the music format is correct and (ii) the twitter login and password are correct for the update.
\end{example}

\begin{figure*}
\center
\includegraphics[width=0.9\textwidth]{ServiceProcess.pdf}
\caption{\label{fig:CIM:serviceprocess} $\pi$-ServiceProcess model for ``To Publish Music''.}
\end{figure*}

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-ServiceComposition} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
provides meta-classes to represent workflows$^{12}$\footnote{$^{12}$.Workflows are transformed into implemented service compositions.} that model  business processes.
The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy.
 The  ($\pi$-Serv\-ice\-Com\-po\-si\-tion) meta-model defines:
%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} meta-class, to represent the classes of entities that collaborate in  business processes by performing some  required action.
An instance of this meta-class is graphically represented as a partition in the activity diagram.
A collaborator can be either internal or external to the system.
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}$^{13}$\footnote{$^{13}$.We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

\item {\sc Action}s, a kind of {\sc ExecutableNode}, are represented in the model as a class activity instance of the meta-class \textsc{Action}.
A class action represents some type of transformation or processing.
There are two types of actions: i) a WebService (attribute Type is {\sf WS}); and ii) a simple operation called an {\sc ActivityOperation} (attribute Type is {\sc AOP}).

>>>>

@@

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

@@

The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.

@@

The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}. In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.

@@
